Exploring environmental impacts of colonial contacts on Indigenous communities in northeastern Taiwan

### 2024-08-06 analysis/paper/paper.qmd
minor edits

# European encounters in the northern region of Taiwan
<!-- the Spanish colonization and the power relations in particular spatial and temporal contexts-->
In 1626, the Spanish East Indies Company (Indias Orientales Españolas) arrived in northern Taiwan and established the San Salvador colony at Heping Dao in Keelung, followed by Santo Domingo in Tamsui, to counter the threat of the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) in the Asian market, which had established a colony in southern Taiwan two years earlier [@Borao2009, p.41; @Andrade2007]. Although the Spanish intended to build a colonial city, San Salvador remains its early development during their colonization, primarily consisting of a fortress and a church [@Borao2009, pp. 103-134; @Berrocal2018]. The neighboring area of the fort was inhavited by local indigenous groups, the Taparris and the Quimaurris, who had lived at the location of the fort before being expelled by the Spanish [@Borao2001, SIT:165, p.81]. The relations between the Spanish and indigenous people might not involve extensive colonialism or intercultural domination, and the Spanish territorial control could be limited to the fort. Instead, the Spanish relied on the exchange of local goods and food surpluses with local indigenous people, according to written records [@Borao2001, SIT:170-171]. This corresponds with a higher proportion of indigenous and Chinese items with little European material found in San Salvador [@Berrocal2018], indicating cultural entanglement.

<!-- the Dutch domination and the processes by which it was established, how and why colonial powers were able to exert control -->
In 1642, the Dutch defeated the Spanish and took over San Salvador and Santo Domingo, expanding their colonial power to northern and eastern Taiwan until 1662. Compared to the Spanish, the Dutch were more strategic in consolidating their power by forming alliances with indigenous villages, implementing political and economic institutions, and conquering villages that were unwilling to comply [@Kang2016; @Andrade2007]. For example, the VOC governors used various colonial institutions to secure their authority, including the annual landdag gathering, taxation, tribute system, village lease system, census investigations, and the selection of indigenous representatives [@Kang2016; @Andrade1997; @Andrade2007]. Among the political institutions, the tribute system operated from 1642 to 1647 and the village lease system introduced in 1644 had great impacts on the local indigenous economy and environment. Based on the tribute system, each household in allied villages was required to pay rice or high-quality deerskin annually. The amount of deerskins collected from indigenous people in northern Taiwan once reached 7,000 pieces in 1646 [@Andrade2007]. Moreover, to increase deerskin exports to Japan, the Dutch implemented a hunting license system (1636-1645) that gave Han Chinese permission to hunt deer on indigenous lands during certain periods [@Heyns2001]. From 1633 to 1661, the average number of deerskins exported from Taiwan was about 67,000 per year [@Koo2011; @Cheng2017]. Those colonial activities accelerated the decline in the deer population and could lead to a harsh situation for the indigenous economy.

<!--the variation of colonial powers over native populations, indirect colonization at Kavalan-->
Although the Dutch power expanded across nearly all of Taiwan in the middle of the 17th century, the extent of their colonial dominance and control differed greatly among different regions. Colonial rules in regions far from the European forts were usually difficult and unsuccessful. Yilan in northeastern Taiwan is an example of minimal European presence and indirect colonization, with fewer imposed colonial rules for most of the time [@Kang2012]. Yilan is an alluvial plain, located about 70 kilometers from the European colony in Keelung, with mountains in between. The main inhabitants in Yilan were the local indigenous group, the Kavalan. Given its strategic location on the route to Torrobouan, where the Dutch believed gold mines existed, they were eager to bring this area under their control [@DZI2000; @Kang2016, p47]. In 1644, the Dutch invaded Yilan to incorporate the defeated Kavalan villages into their northern colonial territory and demanded tribute from the indigenous people [@Kang2016]. Instead of engaging directly with the Kavalan people, the Europeans relied heavily on the assistance of the Basay people active in northern Taiwan to coordinate and collect tributes in Yilan [@Kang2016]. This indirect approach allowed the maintenance of autonomy and flexibility for the Kavalan people, who were able to negotiate tribute requested by the VOC [@Kang2016].

<!-- pre-exiting trade network -->
Before the arrival of the Europeans in northern Taiwan, an inter-insular trade network had been established since the early 15th century, involving Chinese merchants and indigenous groups in Tamsui and Keelung [@Chen2005]. Chinese traders transported goods such as textiles, porcelains, beads, and metal items to exchange for raw materials from Taiwan, such as gold, sulfur, and animal hides and meats from the local Basay people [@Chen2005; @Andrade2007]. This regional trade network was further connected to maritime trade between Ming China and the Ryukyu Kingdom during the 15th and 16th centuries and later to European long-distance trade in the 17th century [@Chen2005; @Von2019; @Wang2007].Yilan was included in this long-established trade network, where the Kavalan people exchanged products such as rice and deerskins for foreign goods and handicrafts from the Basay people [@Kang2003]. This is evidenced by the large amount of imported foreign goods found across archaeological sites in Yilan, indicating frequent trade activities, especially around the 17th century [@TaiwanHistory2005]. Yilan experienced frequent trading activities driven by the pursuit of foreign goods even before the Europeans arrival, and these activities continued during the European presence in northern Taiwan, likely with some occasional pressure from European powers. This unique historical and spatial context makes Yilan an ideal location to study indigenous reactions to foreign influences and the associated trade networks from a long-term perspective in a pericolonial context [@Acabado2017; @Kang2016, pp. 31-78; @Trabert2017].

# Archaeological context at Kiwulan
Kiwulan, located in the northern part of Yilan, is a multi-component site, where the upper component spans from the late Iron Age to the historical periods (AD 1350-1850) [@Chen2007]. As a salvage project, a large scale excavation was conducted at Kiwulan to recover information before it was destroyed. A total of 262 excavation units uncovered abundant artifacts and features, including middens, burials, and post-holes with wooden posts in situ. It suggests that Kiwulan is a well-developed settlement with longhouses (approximately 40 x 80 m) situated side by side and burials located to the north of the residential area [@Chen2004]. The main inhabitants of Kiwulan were the Kavalan people, as indicated by local oral history and the similarities between Kavalan material culture recorded in historical records and archaeological artifacts [@Chen2004]. According to a mid-17th-century Dutch census, there were about 40 indigenous villages with approximately 10,000 people, and Kiwulan (referred to as Kibannoran in Dutch records) was the largest village, with 160 households and 711 people [@Nakamura1938]. In the early 19th century, the population of Kiwulan dramatically decreased to fewer than 100 people due to the population pressure of Han Chinese migrants, which led to the relocation of Kavalan people to eastern Taiwan [@Zhan1995].

The most common artifact was locally made pottery with geometric patterns as surface treatments, which was also found at many simultaneous sites in Yilan. In addition to pottery, a variety of foreign goods, such as porcelains, stonewares, glass beads, carnelian beads, and metal artifacts, were largely found in both the residential area and burials. Due to no evidence of local production of beads or metallurgy, these items are believed to have been acquired through a regional exchange network, primarily with the Basay people active in northern Taiwan. This indicates that Kiwulan has continuously participated in a regional trade network closely associated with northern Taiwan since the 15th century and influenced by the long-distance trade during the 17th century [@]. Moreover, the use of foreign goods appears to have some cultural significance, and previous studies suggested a possible increase in social differentiation due to the introduction of foreign goods [@Wang2020ornament; @Wang2021].

There were a wide range of animal bones and plant remains uncovered, reflecting the diversity of the local diet and the environment close to river and coastal area. The animals include deer, pigs, cows, birds, fish, shellfish, and plants mainly consisting of rice, peach, pandanus, and cucurbits [@Chen2004]. Rice farming seems to be common but limited to a household scale, according to historical records [@]. Shellfish consist of bivalves such as clams and oysters and univalves from rivers and brackish water environments, which indicates a reliance on aquatic and marine resources. Also, those archaeological remains correspond with tools related to subsistence found at Kiwulan, including iron arrowheads for hunting, iron sickles for crop harvesting, and wooden paddles for water-related activities. According to those pieces of evidence, the subsistence practices at Kiwulan should be primarily reliant on rice farming, supplemented by hunting and fishing for wild resources.

>>>>

# European encounters in the northern Taiwan
<!-- Spanish colonization and the power relations in particular spatial and temporal contexts-->
In 1626, the Spanish East Indies Company (Indias Orientales Españolas) arrived in northern Taiwan and established the San Salvador colony at Heping Dao in Keelung, followed by Santo Domingo in Tamsui, to counter the threat of the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) in the Asian market, which had established a colony in southern Taiwan two years earlier [@Borao2009, p.41; @Andrade2007]. Although the Spanish intended to build a colonial city, San Salvador remains its early development during their colonization, primarily consisting of a fortress and a church [@Borao2009, pp. 103-134; @Berrocal2018]. The neighboring area of the fort was inhavited by local indigenous groups, the Taparris and the Quimaurris, who had lived at the location of the fort before being expelled by the Spanish [@Borao2001, SIT:165, p.81]. The relations between the Spanish and indigenous people might not involve extensive colonialism or intercultural domination, and the Spanish territorial control could be limited to the fort. Instead, the Spanish relied on the exchange of local goods and food surpluses with local indigenous people, according to written records [@Borao2001, SIT:170-171]. This corresponds with a higher proportion of indigenous and Chinese items with little European material found in San Salvador [@Berrocal2018], indicating cultural entanglement.

<!-- the Dutch domination, how and why colonial powers were able to exert control -->
In 1642, the Dutch defeated the Spanish and took over San Salvador and Santo Domingo, expanding their colonial power to northern and eastern Taiwan until 1662. Compared to the Spanish, the Dutch were more strategic in consolidating their power by forming alliances with indigenous villages, implementing political and economic institutions, and conquering villages that were unwilling to comply [@Kang2016; @Andrade2007]. For example, the VOC governors used various colonial institutions to secure their authority, including the annual landdag gathering, taxation, tribute system, village lease system, census investigations, and the selection of indigenous representatives [@Kang2016; @Andrade1997; @Andrade2007]. Among the political institutions, the tribute system operated from 1642 to 1647 and the village lease system introduced in 1644 had great impacts on the local indigenous economy and environment. Based on the tribute system, each household in allied villages was required to pay rice or high-quality deerskin annually. The amount of deerskins collected from indigenous people in northern Taiwan once reached 7,000 pieces in 1646 [@Andrade2007]. Moreover, to increase deerskin exports to Japan, the Dutch implemented a hunting license system (1636-1645) that gave Han Chinese permission to hunt deer on indigenous lands during certain periods [@Heyns2001]. From 1633 to 1661, the average number of deerskins exported from Taiwan was about 67,000 per year [@Koo2011; @Cheng2017]. Those colonial activities accelerated the decline in the deer population and could lead to a harsh situation for the indigenous economy.

<!--variation of colonial powers across regions and indirect colonization at Kavalan-->
Although the Dutch power expanded across nearly all of Taiwan in the middle of the 17th century, the extent of their colonial dominance and control differed greatly among different regions. Colonial rules in regions far from the European forts were usually difficult and unsuccessful. Yilan in northeastern Taiwan is an example of minimal European presence and indirect colonization, with fewer imposed colonial rules for most of the time [@Kang2012]. Yilan is an alluvial plain, located about 70 kilometers from the European colony in Keelung, with mountains in between. The main inhabitants in Yilan were the local indigenous group, the Kavalan. Given its strategic location on the route to Torrobouan, where the Dutch believed gold mines existed, they were eager to bring this area under their control [@DZI2000; @Kang2016, p47]. In 1644, the Dutch invaded Yilan to incorporate the defeated Kavalan villages into their northern colonial territory and demanded tribute from the indigenous people [@Kang2016]. Instead of engaging directly with the Kavalan people, the Europeans relied heavily on the assistance of the Basay people active in northern Taiwan to coordinate and collect tributes in Yilan [@Kang2016]. This indirect approach allowed the maintenance of autonomy and flexibility for the Kavalan people, who were able to negotiate tribute requested by the VOC [@Kang2016].

<!-- the pre-existing trade network in norhtern Taiwan-->
Before the arrival of the Europeans in northern Taiwan, an inter-insular trade network had been established since the early 15th century, involving Chinese merchants and indigenous groups in Tamsui and Keelung [@Chen2005]. Chinese traders transported goods such as textiles, porcelains, beads, and metal items to exchange for raw materials from Taiwan, such as gold, sulfur, and animal hides and meats from the local Basay people [@Chen2005; @Andrade2007]. This regional trade network was further connected to maritime trade between Ming China and the Ryukyu Kingdom during the 15th and 16th centuries and later to European long-distance trade in the 17th century [@Chen2005; @Von2019; @Wang2007].Yilan was included in this long-established trade network, where the Kavalan people exchanged products such as rice and deerskins for foreign goods and handicrafts from the Basay people [@Kang2003]. This is evidenced by the large amount of imported foreign goods found across archaeological sites in Yilan, indicating frequent trade activities, especially around the 17th century [@TaiwanHistory2005]. Yilan experienced frequent trading activities driven by the pursuit of foreign goods even before the Europeans arrival, and these activities continued during the European presence in northern Taiwan, likely with some occasional pressure from European powers. This unique historical and spatial context makes Yilan an ideal location to study indigenous reactions to foreign influences and the associated trade networks from a long-term perspective in a pericolonial context [@Acabado2017; @Kang2016, pp. 31-78; @Trabert2017].

# Archaeological context at Kiwulan
<!--introduction of Kiwulan and the population-->
Kiwulan, located in the northern part of Yilan, is a multi-component site, where the upper component spans from the late Iron Age to the historical periods (AD 1350-1850) [@Chen2007]. As a salvage project, a large scale excavation was conducted at Kiwulan to recover information before it was destroyed. A total of 262 excavation units uncovered abundant artifacts and features, including middens, burials, and post-holes with wooden posts in situ. It suggests that Kiwulan is a well-developed settlement with longhouses (approximately 40 x 80 m) situated side by side and burials located to the north of the residential area [@Chen2004]. The main inhabitants of Kiwulan were the Kavalan people, as indicated by local oral history and the similarities between Kavalan material culture recorded in historical records and archaeological artifacts [@Chen2004]. According to a mid-17th-century Dutch census, there were about 40 indigenous villages with approximately 10,000 people, and Kiwulan (referred to as Kibannoran in Dutch records) was the largest village, with 160 households and 711 people [@Nakamura1938]. In the early 19th century, the population of Kiwulan dramatically decreased to fewer than 100 people due to the population pressure of Han Chinese migrants, which led to the relocation of Kavalan people to eastern Taiwan [@Zhan1995].

<!--archaeological materials-->
The most common artifact was locally made pottery with geometric patterns as surface treatments, which was also found at many simultaneous sites in Yilan [@TaiwanHistory2005]. In addition to pottery, a variety of foreign goods, such as porcelains, stonewares, glass beads, carnelian beads, and metal artifacts, were largely found in both the residential area and burials. Due to no evidence of local production of beads or metallurgy, these items are believed to have been acquired through a regional exchange network, primarily with the Basay people active in northern Taiwan. This indicates that Kiwulan has continuously participated in a regional trade network closely associated with northern Taiwan since the 15th century and influenced by the long-distance trade during the 17th century [@]. Moreover, the use of foreign goods appears to have some cultural significance, and previous studies suggested a possible increase in social differentiation due to the introduction of foreign goods [@Wang2020ornament; @Wang2021].

<!--archaeological remains-->
There were a wide range of animal bones and plant remains uncovered, reflecting the diversity of the local diet and the environment close to river and coastal area. The animals include deer, pigs, cows, birds, fish, shellfish, and plants mainly consisting of rice, peach, pandanus, and cucurbits [@Chen2004]. Rice farming seems to be common but limited to a household scale, according to historical records [@]. Shellfish consist of bivalves such as clams and oysters and univalves from rivers and brackish water environments, which indicates a reliance on aquatic and marine resources. Also, those archaeological remains correspond with tools related to subsistence found at Kiwulan, including iron arrowheads for hunting, iron sickles for crop harvesting, and wooden paddles for water-related activities. According to those pieces of evidence, the subsistence practices at Kiwulan should be primarily reliant on rice farming, supplemented by hunting and fishing for wild resources.
